\Large{All in the family:\\Anna and Lisa Hahner's finishing times in\\the 2016 women's Olympic marathon}

### 2017-01-29 paper/paper.tex
some paper edits

@@

\title{Was simultaneous marathon finish in Rio coincidental?}\author{David Cottrell\thanks{Postdoctoral Research Fellow,
Program in Quantitative Social Science, Dartmouth College.  6108
Silsby Hall, Hanover, NH 03755(\texttt{david.cottrell@dartmouth.edu}).} \and Michael C.\
Herron\thanks{Visiting Scholar, Hertie School of Governance, Berlin,
Germany, and Professor of Government, Dartmouth College. 6108 Silsby
Hall, Hanover, NH 03755-3547
(\texttt{michael.c.herron@dartmouth.edu}).}}

@@

On August 14, 2016 at 8:30 in the morning (check) the Women's marathon kicked off in Rio De Janeiro, Brazil, when 156 olympic runners from 80 countries around the world took off from the starting line en route to their destination 42.195 kilometers away.  Two hours, twenty-four minutes, and four seconds later, Jemima Sumgong of Kenya would be the first to cross the finish line and take home gold.  She was just 3.5 minutes behind her personal best.   Approximately 21 minutes later, Germany's twin marathoners,  Anna and Lisa Hahner, would cross the finish line together, holding hands and celebrating a personal victory.  Although they would finish 81st and 82nd respectively and record a time slightly more than 18 minutes behind their personal bests, Anna Hahner would describe the event as a ``magical moment.'' \href{https://www.nytimes.com/2016/08/17/sports/olympics/twins-finish-marathon-hand-in-hand-but-their-country-says-they-crossed-a-line.html}   The media quickly picked up on the story as an image of the two twins finishing hand-in-hand with beaming smiles captured a public audience.  Many believed that the moment was a reflection of the Olympic spirit.

However, not everyone agreed.  The sports director of the German Athletics Federation, Thomas Kurschilgen, immediately stirred up controversy when he suggested that the twins' photo-finish was no coincidence.  He stated that their main goal was to create a spectacle and ``generate media attention,'' using the fact that they were at least 18 minutes behind their personal best times as evidence that they were slowing their pace in order to finish simultaneously.\footnote{\url{https://www.nytimes.com/2016/08/17/sports/olympics/twins-finish-marathon-hand-in-hand-but-their-country-says-they-crossed-a-line.html}}    Of course, these accusations were denied by Anna and Lisa Hahner, who claimed that their simultaneous finish was simply an unintended coincidence.

We wondered what the data said about this likelihood.  Did Anna and Lisa Hahner finish the race together by chance or was it something more calculated?

@@

Thomas Kurschilgen's accusations of an intentional finish stems from two observations about the race.  First, the twins finished slower than he expected, recording a time that is more than 18 minutes behind their personal best.  Second, the twins finished together at the exact same time, which seemed like an unlikely coincidence.   Kurschilgen clearly believed that neither of these events would have occurred had Anna and Lisa run the marathon independently, absent any coordination. Instead, the twins' would have run at a faster pace and they would not have finished simultaneously.

On the other hand, only a handful of runners completed the marathon in Rio with times that were faster than their recorded best time. Therefore, 18 minutes behind their personal record may not be the outlier that Kurschilgen claimed it to be.  Moreover, there is good reason to suspect that an unintended simultaneous finish is more likely for Anna and Lisa than it would be for any other set of runners. After all, they are twins with similar abilities.  Not only do they train together, but their record best times are less than two minutes apart.  While Anna may be slightly faster than Lisa, given their personal bests, we might expect the difference in their result to be just as close as the difference in the their recorded bests.   And given random variation in finishing - some days are better than others - a simultaneous finish might not be out of the ordinary.

To test the assertion that Anna and Lisa Hahner intentionally slowed their pace to finish simultaneously, we simply need to compute the probability that such a result would have occurred unintentionally.  Hence we need to know the probability distribution of Anna and Lisa's finishing times given that their runs were completely independent of each other.    The challenge, of course, is that this distribution is unknown.  Therefore it needs to be estimated.

@@

We know that Anna and Lisa Hahner finished the marathon in Rio simultaneously.  We don't know, however, if the finish was intentional; either the twins acted independently $I$ or they acted in coordination $C$.  These are the two possible states of the race $\psi = \{I, C\}$.  Given that we observed Anna and Lisa's final times, $Y_A =  y_A$ and $Y_L = y_L$, we want to know the probability that they ran independently, as they say they did. Hence, we are looking to determine,

@@

However, to determine this, we need to have some understanding of the likelihood function.  We want to know the likelihood of Anna's and Lisa's final times given independence $P(y_A  \cap y_L \mid \psi = I )$.  
We can estimate this function by beginning with a few assumptions.  First, we assume that under independence, any given runner's final time $Y_i$ is conditional on his/her running ability plus noise.   Specifically, we assume that the $Y_i$ is a linear function of the runner's ability $X_i$ plus a normally distributed error term $e_i \sim N(0, \sigma)$.   Second, we assume that every runner shares the same linear relationship - meaning the slope and intercept remain constant across runners.   Third, we assume that the error term is constant across runners as well.  Hence, luck and misfortune are drawn from the same distribution.  Therefore,

$$Y_i \sim N(\beta_0 + X_{i}\beta_1 + e, \sigma)$$

We also assume that a runner's ability $X_i$ can be measured precisely by their best marathon performance leading up to the Olympics. Any measurement error must therefore be negligible.

@@

\section*{Did Anna and Lisa intentionally slow down?}
According to Kurschilgen, Anna and Lisa underperformed in the Rio marathon. He claimed that because their goal was to finish simultaneously rather than finish at their fastest pace, their times were slower than they otherwise would have been.  He claimed that the twins were simply trading speed for a photo-finish.

@@

\begin{figure}[!ht]
 \caption{Relationship between Personal Best and Result}
 \label{fig:scatter}
 \begin{subfigure}{.5\textwidth}
 \includegraphics[width=\textwidth, keepaspectratio]{scatter_plot.pdf}
 \end{subfigure}
 \begin{subfigure}{.5\textwidth}
 %\label{fig:Studentized Residuals}
 \includegraphics[width=.975\textwidth, keepaspectratio]{studentized_residuals.pdf}
 \end{subfigure}
\end{figure}

>>>>

\newcommand{\yi}{\ensuremath{Y_i}}

@@

\title{\Large{All in the family:\\Anna and Lisa Hahner's finishing
    times in\\the 2016 women's Olympic
  marathon}}\author{David Cottrell\thanks{Postdoctoral Research
  Fellow, Program in Quantitative Social Science, Dartmouth College,
    6108 Silsby Hall, Hanover, NH
    03755 (\texttt{david.cottrell@dartmouth.edu}).} \and Michael C.\
  Herron\thanks{Visiting Scholar, Hertie School of Governance, Berlin,
    Germany, and Professor of Government, Dartmouth College, 6108
    Silsby Hall, Hanover, NH 03755
    (\texttt{michael.c.herron@dartmouth.edu}).}}

@@

On August 14, 2016, at 9:30 in the morning, the Women's Olympic
marathon kicked off in Rio de Janeiro, Brazil, when 156 runners from
80 countries across the world took off from the starting line en route
to their destination 42.195 kilometers away. Two hours, twenty-four
minutes, and four seconds later, Jemima Sumgong of Kenya would be the
first to cross the finishline and take home gold; Sumgong was just 3.5
minutes behind her personal best time in the marathon. Approximately
21 minutes later, twin marathoners from Germany, Anna and Lisa Hahner,
would cross the finishline together, holding hands and celebrating a
personal victory. Although the Hahners would finish 81st and 82nd,
respectively, with times slightly more than 18 minutes slower than
corresponding personal bests, Anna Hahner would describe their joint
finish as a ``magical moment.''
\href{https://www.nytimes.com/2016/08/17/sports/olympics/twins-finish-marathon-hand-in-hand-but-their-country-says-they-crossed-a-line.html}
The media quickly picked up on the Hanher story as an image of the
beaming twins finishing hand-in-hand captured a public audience. Many
believed the moment was a reflection of the Olympic spirit.

Not everyone agreed with this rosy interpretation. The twins' happy
facial expressions at the finish were portrayed as a bit contrived
(smiling like ``Honigkuchenpferde''---a Honigkuchenpferd is a German
cookie in the shape of a horse), and the sports director of the German
Athletics Federation, Thomas Kurschilgen, stirred up controversy when
he suggested that the Hahners' photo-finish was no coincidence.
Kurschilgen averred that the twins slowed down so as to finish
simultaneously and sought to create a spectacle and ``generate media
attention.'' Kurschilgen justified his charge with the fact that the
twins were at least 18 minutes behind their personal best
times.\footnote{\url{https://www.nytimes.com/2016/08/17/sports/olympics/twins-finish-marathon-hand-in-hand-but-their-country-says-they-crossed-a-line.html}}
Not surprisingly, these accusations were denied by the Hahners, who
claimed that their simultaneous finish was simply an unintended
coincidence.

%% Honigkuchenpferd quote from here:
%%
%% https://www.welt.de/sport/olympia/article157669264/Das-falsche-Laecheln-der-deutschen-Lauf-Zwillinge.html

What happened in the women's Olympic marathon, and how might we assess
whether the Hahner finish was coincidental or intentional? These two
interpretations are clearly at odds. If the former, then the Hahner
twins are to be celebrated and their finish treated as an expression
of the spirit behind the Olympic games. If the latter, though, then
the Hahner twins may have violated this spirt by not trying hard
enough. It is perhaps too easy for us to write such a glib
sentence---neither of us cannot fathom being able to complete a
marathon anywhere in the vicinity of two and a half hours---but we
nonetheless want to know what the data from the Olympic marathon data
tell us. Was the women's Olympic marathon a lovely coincidence or
something else?

Among female Olympic marathoners, the Hahner twins were not alone in
their familial ties, and we draw on this in the analysis that follows.
The marathon also featured twins from North Korea, Kim Hye-song and
Kim Hye-gyong, who had identical times and finished 10th and 11th,
respectively. The Kim finish, unlike the Hahner finish, appears devoid
of post-race controversy. Moreover, three triplets from Estonia
competed in the Rio marathon, although only two, Lily Luik and Leila
Luik, finished it, in 97th and 114th place, respectively. The third
Estonia triplet, Liina Luik, recorded what is known as a DNF---an
abbreviation that means did not finish, a term that we will use
throughout this article.

\section*{Marathon data}

For each participant who started the women's Olympic marathon, we know
several things: personal best marathon time prior to the 2016 Olympic
games; split times from the Rio marathon course at 5 kilometers, 10
kilometers, and so forth; and, finishing time. We cannot directly
observe the effort that an individual put into the race, and we do not
know why some runners have DNF results; some runners may have injured
themselves on the course and accordingly dropped out, and others may
have dropped out, uninjured, in anticipation of an unsatisfactory
result. Of the 156 marathon starters, 133 completed the race and 23
DNFed at various locations throughout the course. The overall DNF rate
was thus $\frac{23}{156} \approx 0.15$, and the relatively small
sample size at our disposal---156 runners---means that a 95\%
confidence for this rate is fairly wide, namely,
$\left(0.098, 0.22\right)$.

Key in the analysis that follows is the deviation between a runner's
Rio time and her prior personal best time in the marathon. Of course
there is variance across runners in these deviations, and we want to
know if patterns in the deviations support an honest effort by the
Hanher twins or perhaps something else.

@@

Thomas Kurschilgen's accusations of an intentional finish stems from
two observations. First, the Hahner twins finished slower than he
expected, as we already noted recording times more than 18 minutes
slower than respective personal bests. Second, the twins finished
together at the exactly same time. Kurschilgen clearly believed that
neither of these events would have occurred had both Hahner run the
marathon independently, absent coordination. By his logic, the German
twins should have run faster and not have finished simultaneously.

On the other hand, only a handful of runners completed the Rio
marathon with times that were faster than their recorded best times.
Thus, 18 minutes behind a personal record may not be the outlier that
Kurschilgen claimed it to be. Moreover, there is good reason to
suspect that an unintended simultaneous finish was more likely for the
Hahner twin than it would be for any other set of runners. After all,
these two women are twins with presumably similar abilities. Not only
do they train together, but their pre-Rio best times are less than two
minutes apart. While Anna may be slightly faster than Lisa, measured
by personal bests, we might expect the difference in their Rio result
to be just as close as the differences in their recorded bests. And
given random variation in finishing times, a simultaneous finish might
not be out of the ordinary.

Even though they seem similar, finishing at very similar times and
finishing together are different phenomena. If, for example, both
Hahners were of similar ability to each other and also to many other
runners, then we might expect similar finishing times yet not
necessarily similar placements. The latter will be a function of the
extent to which all runners on the Rio course have similar talent
levels.  This point is an important one and will be evident in the
results that follow.

To test the assertion that the Hahner twins paced themselves to finish
at the same time and with back-to-back placements, we simply need to
compute the probability that such a result would have occurred
unintentionally. Hence we need to know the probability distribution of
Anna and Lisa's finishing times if their runs had been independent of
each other. The challenge, of course, is that this distribution is
unknown.  ^^ Therefore it needs to be estimated.

@@

%% Ok.  Not sure what all the notation means.

% We know that Anna and Lisa Hahner finished the marathon in Rio
% simultaneously. We don't know, however, if the finish was intentional;

Suppose that Hahner the twins acted independently $I$ or they acted in
coordination $C$; these are the two possible states of the race
$\psi = \{I, C\}$. Given that we observed Anna and Lisa's finishing
times, $Y_A = y_A$ and $Y_L = y_L$, we want to know the probability
that they ran independently, as they say they did. Hence, we are
looking to determine,  %% y_a versus Y_a.  Former realized?

@@

However, to determine this, we need to have some understanding of a
likelihood function that specifies Rio finishing times. We want to
know the likelihood of Anna's and Lisa's final times given
independence $P(y_A \cap y_L \mid \psi = I )$. We can estimate this
function with a few assumptions. First, we assume that under
independence, any given runner's final time $Y_i$ is conditional on
his/her running ability plus noise. Specifically, we assume that the
$Y_i$ is a linear function of the runner's ability $X_i$ plus a
normally distributed error term $e_i \sim N(0, \sigma)$. Second, we
assume that every runner shares the same linear relationship - meaning
the slope and intercept remain constant across runners. Third, we
assume that the error term is drawn from a common distribution across
runners. Hence, luck and misfortune are drawn from the same
distribution. Therefore,

$$Y_i \sim N(\beta_0 + X_{i}\beta_1 + e_i, \sigma)$$

We also assume that a runner's ability $X_i$ can be measured precisely
by her best marathon performance leading up to the Olympics.

%% Any measurement error must therefore be negligible.

@@

\section*{Did Anna and Lisa intentionally slow down?}

According to Kurschilgen, Anna and Lisa underperformed in the Rio marathon. He claimed that because their goal was to finish simultaneously rather than finish at their fastest pace, their times were slower than they otherwise would have been.  He claimed that the twins were simply trading speed for a photo-finish.

@@

\begin{figure}[!ht]
  \caption{Relationship between Personal Best and Result  -- better label?}
  \label{fig:scatter}
  \begin{subfigure}{.5\textwidth}
    \includegraphics[width=\textwidth,
    keepaspectratio]{scatter_plot.pdf}
    \caption{Rio finishing times and personal best times}
    \label{fig:A}
  \end{subfigure}
  \begin{subfigure}{.5\textwidth}
    % \label{fig:Studentized Residuals}
    \includegraphics[width=.975\textwidth, keepaspectratio]{studentized_residuals.pdf}
    \caption{A}
    \label{fig:B}
  \end{subfigure}
\end{figure}
